Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-12 numerical.tex
rough revision done

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode according
to explicit solutions developed in Appendix \ref{iso_explicit}. We
show a meridional visualization of this mode in
Fig. \ref{lowfreq_eigenfunc_2d}. Although there is a constant vertical
velocity throughout the fluid column, most of the momentum is
contained within $\sim 2H$ of the midplane due to the density
stratification.

@@

\subsubsection{Surface modes: $\khat=30$} 
Our standard setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (red crosses) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes' in
our case. These modes lie parallel to the vertical axis in mode
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for
sufficiently large $\zmax$ and/or $\khat$. % For example, the two
% largest growth rates found for the case with $\zmax=7H$ in
% Fig. \ref{compare_modes_iso_kx10} are surface modes.

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes can be seen with
$\zmax\geq 5H$, examples of which are marked by `s'. These modes have
been discussed in the above studies. They typically have the largest
growth rates but require sufficiently large vertical domain size and their eigenvalues
vary with $\zmax$.

@@

\subsection{Effect of thermal relaxation}
We now consider the effect of thermal relaxation by increasing
$\beta$. Note that for the standard disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. This suggests that tracking fundamental mode would be a
more robust way of assessing stability as a function of $\beta$,
rather than searching for the most 
unstable mode, because the latter may be associated with boundary
conditions. In fact, the fundamental mode bcomes dominant 
(or close to it) with increasing $\beta$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_beta0d1}
  \caption{Same as Fig. \ref{lowfreq_eigenfunc_cool} but with a
    thermal relaxation timescale $\beta=0.1$. 
    \label{lowfreq_eigenfunc_cool}
  }
\end{figure}

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI as a function of $\beta$ for $\khat\in[1,10^2]$. We
also plot the critical thermal relaxation timescale
$\beta_\mathrm{crit}=0.125$ given by Eq. \ref{iso_vsi_cond} for the present
disk parameters.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides an
accurate prediction of the upper limit to the thermal relaxation 
timescale for  the fundamental mode. For smaller $\khat$,
$\beta<\beta_\mathrm{crit}$ is a sufficient condition for
instability. For larger $\khat$, we find $\nu$ does not reach zero but
rapidly decreases as $\beta$ approaches and increases beyond
$\beta_\mathrm{crit}$. In the latter case we find the perturbation
acquires large amplitudes near the vertical boundaries. Boundary
conditions may thus may play a role for these modes that is not
accounted by our analytical discussion.

Fig. \ref{bcrit_compare1} shows that 
$\beta_\mathrm{crit}$ is a good indication for the maximum
thermal relaxation timescale to permit the fundamental VSI. For
$\beta\lesssim O(10^{-2})$, growth rates are insensitive to thermal
relaxation. For this disk model, \citeauthor{nelson13}
found a thermal relaxation timescale of $\beta\gtrsim 0.6$ stabilized
the VSI in non-linear hydrodynamic simulations. This is consistent
with our results.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$  obtained
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental VSI on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately. Here, we use a wavenumber $\khat=10$, for which we
find growth rates reach zero for the range of disk parameters
considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
$\gamma$, i.e. weaker instability, although there is a noticeable difference
as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
assumed $\gamma>1$.

Fig. \ref{bcrit_compare}, together with the previous section
(Fig. \ref{bcrit_compare1}), shows that $\beta_\mathrm{crit}$ as given
by Eq. \ref{iso_vsi_cond} gives a measure of the characteristic
thermal timescale below which perturbations are effectively isothermal
so that the fundamental VSI operates.

@@

\subsection{Vertically non-isothermal disks} 
We briefly consider vertically non-isothermal disks with 
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax\simeq2.2H$, which
is close to the zero-density surface.

@@

% \subsection{VSI in a stably stratified disk}
% We demonstrate the (strong) stabilizing effect of a  
% positive vertical entropy gradient ($N_z^2\geq0$) by setting $\gamma=2$.  
% Since we expect the perturbations to decay rapidly
% away from the midplane (see \S\ref{analytic_adia}), we consider a
% smaller domain with $\zmax = 2\epsilon r$. 
% Other disk and perturbation parameters are the same as in
% \S\ref{vertiso_pertiso}.

% Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
% problem. The growth rates are much smaller than
% those in the previous section. The fundamental VSI mode is that with
% $\mathrm{max}\nu$. Its expected and numerically-calculated growth
% rates compares well, with 
% \begin{align*}
%   \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
%     Eq. \ref{gam2_growth_rate})},\\
%   \nu = 0.03672 \epsilon \Omega_k &\quad \text{(numerical)}.
% \end{align*}
% The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
% and clearly show that perturbations are rapidly stabilized away from
% the midplane. For the adopted disk and perturbation parameters,
% Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$, corresponding
% to a characteristic decay length scale of $\sqrt{2/|\alpha|}\epsilon r\simeq 0.2\epsilon r$,
% which is consistent with that observed in  Fig. \ref{lowfreq_eigenfunc_adia}.

% This decay lengthscale is rougly the height within which
% vertical shear is larger than the buoyancy frequency squared. That is,
% \begin{align*}
% \left|r\frac{d\Omega^2}{dz}\right|\gtrsim N_z^2
% \end{align*}
% for
% \begin{align*}
%   \left|\frac{z}{\epsilon r}\right| \lesssim \frac{\epsilon|q|\gamma}{\gamma-1},
% \end{align*}
% in a thin, vertically isothermal disk. This evaluates to $|z|\lesssim
% 0.2\epsilon r$ for the current disk model, consistent with
% Fig. \ref{lowfreq_eigenfunc_adia}.

% Furthermore, the growth rate to should be no larger than the
% maximum vertical shear available within this decay lengthscale. Using 
% Eq. \ref{max_growth} as a rough guide, 
% \begin{align}
%   \frac{\nu}{\epsilon\Omega_k}<\frac{\epsilon |q|^2\gamma}{2(\gamma-1)},
% \end{align}
% or $\nu< 0.1\epsilon\Omega_k$ in our disk model,
% consistent with numerical results.

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/eigenvalues_adia}
%   \caption{Eigenvalues in for the
%     VSI in a nearly vertically isothermal disk
%     ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
%     evolved adiabatically with $\gamma=2$. The perturbation radial
%     wavenumber is $\khat=20\pi$. \label{lowfreq_eigen_adia}
%   }
% \end{figure}

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/eigenvectorvz_adia}
%   \caption{Fundamental VSI mode in a nearly vertically
%     isothermal disk ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
%     evolved adiabatically
%     with $\gamma=2$. This eigenfunction 
%     corresponds to the top-left eigenvalue displayed in 
%     Fig. \ref{lowfreq_eigen_adia} (largest $|\nu|$).  
%     The  real (imaginary) part of $\delta v_z$ are shown as solid
%     (dotted) lines. 
%     \label{lowfreq_eigenfunc_adia}
%   }
% \end{figure}

% Although a positive vertical entropy gradient is stabilizing, we note
% that $N_z^2\propto z^2$  while vertical shear $rd\Omega^2/dz\propto
% z$, so that there is always a region sufficiently close to the
% midplane in which vertical shear dominates over bouyancy and the VSI
% can operate. However, the growth rates are small and the instability
% only affects the regions close the midplane. These modes may not be
% important in practice.

% \subsection{Other adiabatic indices}
% To see the transition from neutrally-bouyant to stably stratified cases above, 
% we repeated the previous calculation with $\gamma\in[1,2]$. 
% We restore the vertical domain to $\zmax=10\epsilon r$. 
% Fig. \ref{adia_growth_num} shows the eigenfrequencies of the
% fundamental VSI mode.

% %To select the appropriate
% %eigenvalues we vary the parameters slowly away from a reference solution.   
% % The numerically-determined eigenfrequencies are qualitatively
% % consistent with the discussion in \S\ref{adia_compare_gam1}.
% There is good agreement between Fig. \ref{adia_growth_num} and
% Fig. \ref{adia_growth} for $\khat=k_xH\gtrsim1$. There is a
% mismatch for smaller wavenumbers, especially in the growth
% rates. Actual growth rates decrease slower as $\khat\to0$
% than that in Fig. \ref{adia_growth}. Disagreement at small
% $\khat$ is not unexpected because the low-frequency
% approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
% $|\omega|\sim\Omega_k$ for small $\khat$.

% Given the number of approximations made in \S\ref{adia_compare_gam1},
% Fig. \ref{adia_growth_num} compares well with 
% Fig. \ref{adia_growth}. The main qualitative features of the full
% numerical solution in Fig. \ref{adia_growth_num} are captured by
% Fig. \ref{adia_growth}: stabilization for 
% $\khat\to0$ and $\khat\to\infty$,
% stabilization for increasing $\gamma$ (but ineffective at small $\khat$), 
% and the shift of the most unstable wavenumber to smaller values as
% $\gamma$ is increased.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
%   \caption{Real frequency (top) and growth rate 
%     (bottom) of the fundamental VSI in a nearly vertically   
%     isothermal disks ($\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$, 
%     subject to adiabatic perturbations with different values of $\gamma$. 
%     The eigenfrequencies are calculated by numerically
%     solving the full linear eigenvalue problem. This plot is to
%     be compared with Fig. \ref{adia_growth}. \label{adia_growth_num}}   
% \end{figure}

>>>>

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode according
to explicit solutions developed in Appendix \ref{iso_explicit}. We
show a meridional visualization of this mode in
Fig. \ref{lowfreq_eigenfunc_2d}. Radial velocities are typically much
smaller than vertical velocities. Although there is a constant vertical
velocity throughout the fluid column, most of the momentum is
contained within $\sim 2H$ of the midplane due to the density
stratification.

@@

\subsubsection{Comment on `surface modes'} 
Our standard setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (red crosses) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes' in
our case. These modes lie parallel to the vertical axis in mode
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$. % For example, the two
% largest growth rates found for the case with $\zmax=7H$ in
% Fig. \ref{compare_modes_iso_kx10} are surface modes.

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes can be seen with
$\zmax\geq 5H$, examples of which are marked by `s'. These modes have
been discussed in the aforementioned studies. They typically have the
largest growth rates but require sufficiently large vertical domain
size and their eigenvalues vary with $\zmax$.

@@

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of thermal relaxation by increasing
$\beta$. Note that for the standard disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

@@

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. Thus, tracking fundamental mode is a
more robust way of assessing stability as a function of $\beta$,
rather than searching for the most 
unstable mode, because the latter may be associated with boundary
conditions. In fact, the fundamental mode bcomes dominant 
(or close to it) with increasing $\beta$.

Fig. \ref{lowfreq_eigenfunc_cool}---\ref{lowfreq_eigenfunc_2d_cool}
shows the fundamental mode with $\khat=10$ in a disk with 
$\beta=0.1$. These plots can be compared the case with rapid thermal
relaxation (Fig. \ref{lowfreq_eigenfunc}---\ref{lowfreq_eigenfunc_2d}). We see 
that increasing $\beta$ causes the region characterized by $W\sim z$ and  
$\delta v_z\sim$ constant closer to the mid-plane, with oscillatory
behavior emerging near vertical boundaries where buoyancy becomes
important. This leads to the `tilted' pressure perturbations seen in
Fig. \ref{lowfreq_eigenfunc_2d_cool}
(cf. Fig. \ref{lowfreq_eigenfunc_2d}). However, most of the meridional
momentum is still contained within $|z|\lesssim 2H$ as for 
the case with $\beta=10^{-3}$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_beta0d1}
  \caption{Same as Fig. \ref{lowfreq_eigenfunc} but with a
    thermal relaxation timescale $\beta=0.1$. 
    \label{lowfreq_eigenfunc_cool}
  }
\end{figure}

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
We now test the critical thermal relaxation timescale given by
Eq. \ref{iso_vsi_cond}. We follow the fundamental mode by starting
from eigenvalues for $\beta=10^{-3}$ and slowly increase the thermal
relaxation timescale up to $\beta=10^3$. Fig. \ref{bcrit_compare1}
show the obtained growth rates, together with the critical 
thermal relaxation timescale $\beta_\mathrm{crit}=0.125$ for our
fiducial disk parameters.

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, 
$\beta<\beta_\mathrm{crit}$ the VSI can operate slightly beyond
$\beta_\mathrm{crit}$ (recall that $\beta_\mathrm{crit}$ was derived
by assuming $\khat\gg 1$). For $\khat\geq 50$, we find perturbations
acquire large amplitudes near the vertical boundaries. Complications
% to our simple solution procedure 
may arise from surface modes. Nevertheless, growth rates rapidly
decrease as $\beta$ approach and increase beyond
$\beta_\mathrm{crit}$.

Fig. \ref{bcrit_compare1} shows that $\beta_\mathrm{crit}$ provides a  
practical indication for the maximum thermal relaxation timescale to
permit the fundamental VSI. % From our discussion in 
% \S\ref{therm_relax_eff}, which indicate the fundamental mode typically
% dominant 
\cite{nelson13} found a thermal relaxation timescale of $\beta\gtrsim
0.6$ stabilized the VSI in non-linear hydrodynamic simulations. This 
is consistent with our results.

@@

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI in the fiducial disk
     model as a function of the thermal relaxation timescale $\beta$
     The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$  obtained
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We further test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately. Here, we use a wavenumber $\khat=10$, for which we
find growth rates reach zero for the range of disk parameters
considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

The numerically obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
$\gamma$, i.e. weaker instability. There is a noticeable difference
as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
assumed $\gamma>1$.

Fig. \ref{bcrit_compare}, together with the previous section
(Fig. \ref{bcrit_compare1}), shows that $\beta_\mathrm{crit}$ as given
by Eq. \ref{iso_vsi_cond} characterizes the thermal timescale below
which perturbations are effectively isothermal 
so that the VSI operates.

@@

\subsection{Vertically non-isothermal disks} 
We briefly consider vertically non-isothermal disks with 
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax=0.99H_s$.

@@
